---
title: "C01 Games EDA"
author: "Connor Quin"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)

# Parameters
rds_path <- here::here("data/games.rds")

eco_levels = c("A", "B", "C", "D", "E")
#===============================================================================

```


Here is the first draft of my initial C01 analysis on lichess's database chess data. As a first run, I'm using data from Lichess's January 2013 database. The raw PGN file is 16.1 MB, with 121,332 games.

To start, let's read in this data and perform any EDA modifications

```{r}
games <-
  read_rds(rds_path)

summary(games)
```


First off, let's do some 1D visualizations to get a sense of how the continuous variables look


```{r}
# Distribution of white ratings
games %>% 
  ggplot(mapping = aes(x = white_elo)) +
  geom_histogram(binwidth = 10) +
  geom_vline(aes(xintercept = median(games$white_elo), color = "Median"))

# Distribution of changes to white ratings
games %>% 
  ggplot(mapping = aes(x = white_elo_diff)) +
  geom_histogram(binwidth = 10) +
  geom_vline(aes(xintercept = median(games$white_elo_diff), color = "Median"))

# Distribution of black ratings
games %>% 
  ggplot(mapping = aes(x = black_elo)) +
  geom_histogram(binwidth = 10) +
  geom_vline(aes(xintercept = median(games$white_elo), color = "Median"))

```

Some thoughts:

* Note the odd spike at `white_elo = 1500`. This is because Lichess assigns newly signed-up members an initial rating of 1500

* As a median, this dataset saw the white players slightly higher than their black counterparts.


Now for the discrete variables:

```{r}
# Frequency of time types:
games %>% 
  ggplot(mapping = aes(x = time_type)) +
  geom_bar()

# Frequency of results:
games %>% 
  ggplot(mapping = aes(x = result)) +
  geom_bar()

# Frequency of termination:
games %>% 
  ggplot(mapping = aes(x = termination)) +
  geom_bar()

# Opening Frequencies:
games %>%
  transmute(
    eco_start = factor(substring(eco, 1, 1), levels = eco_levels)
  ) %>% 
  ggplot(mapping = aes(x = eco_start)) +
  geom_bar()
```

Thoughts:

* The game has a good mix of bullet, blitz, and classical game. Correspondence is much less represented (n = 180) and might as well be removed later perhaps.

* Compared to top level over-the-board chess, there are very few draws here. White also wins more often than black.

* Most games are terminated on the board, but a healthy amount are also won/lost on time.

* Across all these games, C is the most common ECO code, followed by B, A, D, and E.




